---
title: "Data carpentry"
author: "Robin Lovelace, University of Leeds, \n `r Sys.Date()`"
date: "Slides: [rpubs.com/RobinLovelace](http://rpubs.com/RobinLovelace)"
output:
  ioslides_presentation:
    widescreen: true
    transition: slower
    logo: ../figures/its-logo-square.png
---

## Why data carpentry?

<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr">Data analysts and &#39;scientists&#39;: don&#39;t wrangle, munge or &#39;hack&#39; your valuable datasets. Use <a href="https://twitter.com/hashtag/datacarpentry?src=hash">#datacarpentry</a>: <a href="https://t.co/gXrlIJH91R">https://t.co/gXrlIJH91R</a> <a href="https://t.co/GSWS7O7zBz">pic.twitter.com/GSWS7O7zBz</a></p>&mdash; Robin Lovelace (@robinlovelace) <a href="https://twitter.com/robinlovelace/status/833622374037721089">February 20, 2017</a></blockquote>
<script async src="//platform.twitter.com/widgets.js" charset="utf-8"></script>

- If you 'hack' or 'munge' data, it won't scale
- So ultimately it's about being able to handle Big Data
- We'll cover the basics of data frames and tibbles
- And the basics of **dplyr**, an excellent package for data carpentry
    - **dplyr** is also compatible with the **sf** package

## The data frame

The humble data frame is at the heart of most analysis projects:

```{r}
d = data.frame(x = 1:3, y = c("A", "B", "C"))
d
```

In reality this is a list, making function work on each column:

```{r}
summary(d)
plot(d)
```

## Subsetting

In base R, there are many ways to subset:

```{r}
d[1,] # the first line
d[,1] # the first column
d$x # the first column
d [1] # the first column, as a data frame
```

## The tibble

Recently the data frame has been extended:

```{r}
library("tibble")
dt = tibble(x = 1:3, y = c("A", "B", "C"))
dt
```

## Advantages of the tibble

It comes down to efficiency and usability

 - When printed, the tibble diff reports class
 - Character vectors are not coerced into factors
 - When printing a tibble diff to screen, only the first ten rows are displayed
 
## dplyr

Like tibbles, has advantages over historic ways of doing things

- Type stability (data frame in, data frame out)
- Consistent functions - functions not `[` do everything
- Piping make complex operations easy

```{r, eval=FALSE}
ghg_ems %>%
  filter(!grepl("World|Europe", Country)) %>% 
  group_by(Country) %>% 
  summarise(Mean = mean(Transportation),
            Growth = diff(range(Transportation))) %>%
  top_n(3, Growth) %>%
  arrange(desc(Growth))
```

```{r}
# dplyr must be loaded with
library(dplyr)
```


## Why pipes?

```{r, eval=FALSE}
wb_ineq %>% 
  filter(grepl("g", Country)) %>%
  group_by(Year) %>%
  summarise(gini = mean(gini, na.rm  = TRUE)) %>%
  arrange(desc(gini)) %>%
  top_n(n = 5)
```

vs

```{r, eval=FALSE}
top_n(
  arrange(
    summarise(
      group_by(
        filter(wb_ineq, grepl("g", Country)),
        Year),
      gini = mean(gini, na.rm  = TRUE)),
    desc(gini)),
  n = 5)
```

## Subsetting with dplyr

Only 1 way to do it, making life simpler:

```{r}
select(dt, x) # select columns
slice(dt, 2) # 'slice' rows
```


## Practical prerequisites

You'll need the `efficient` package installed:

```{r, eval=FALSE}
devtools::install_github("csgillespie/efficient")
```


These are the packages we'll be using:

```{r, eval=FALSE}
library("tibble")
library("tidyr")
library("stringr")
library("readr")
library("dplyr")
```

## Practical

- Test out tibbles with [ERP 6.2](https://csgillespie.github.io/efficientR/data-carpentry.html#efficient-data-frames-with-tibble) (15 minutes)
- Tidy data is vital for good data carpentry. Work through the tidying code in section [6.3](https://csgillespie.github.io/efficientR/data-carpentry.html#tidying-data-with-tidyr-and-regular-expressions) (30 minutes)
- Work through in detail [Section 6.4 on dplyr](https://csgillespie.github.io/efficientR/data-carpentry.html#dplyr)
- Work through the code examples in Efficient R Programming (45 minutes)
- Bonus (if time allows): look at how to merge datasets with `left_join()` in [Section 6.5](https://csgillespie.github.io/efficientR/data-carpentry.html#combining-datasets) (15 minutes)
