---
title: "Data carpentry for sustainable transport planning: the Propensity to Cycle Tool"
author: "Robin Lovelace, University of Leeds, \n `r Sys.Date()`"
date: "Slides: [rpubs.com/RobinLovelace](http://rpubs.com/RobinLovelace)"
output:
  ioslides_presentation:
    widescreen: true
    transition: slower
    logo: ../figures/its-logo-square.png
    keep_md: true
---

## The nature of transport data

- The same as handling any other kind of data in R
- But: dynamic, complex (e.g. GTFS) reliant on official feeds
- That's why using a language that is powerful and flexible is good

## Worked example: pct data in West Yorksire

- We'll download and visualise some transport data

```{r}
# u_pct = "https://github.com/npct/pct-data/raw/master/west-yorkshire/l.Rds"
# download.file(u_pct, "l.Rds")
library(stplanr)
l = readRDS("l.Rds")
plot(l)
```

## Analysing where people walk

```{r}
sel_walk = l$foot > 9
l_walk = l[sel_walk,]
plot(l)
plot(l_walk, add = T, col = "red")

library(dplyr) # for next slide...
```

## Doing it with sf (!)

```{r, eval=FALSE}
l_walk1 = l %>% filter(All > 10) # fails
```

```{r}
library(sf)
l_sf = st_as_sf(l)
plot(l_sf[6])
```

## Subsetting with sf | is really fun!

```{r}
l_walk2 = l_sf %>% 
  filter(foot > 9)
plot(l_sf$geometry)
plot(l_walk2$geometry, add = T, col = "green")
```





## Exercises | Use sf or sp depending on what you want

- Find and plot all desire lines where cycling is more common than walking 
- Find all desire which are less than 2.5 km 
- Find all lines in which driving is more common than walking AND are 
- Find and plot all desire lines that are longer. Where are they?
- Use the following code to download bus stops in the area

## Solution

```{r}
l_sf$distsf = as.numeric(st_length(l_sf))
summary(l_sf$distsf)
l_drive_short = l_sf %>% 
  filter(distsf < 1000 & car_driver > foot)
l_drive_short2 = l_sf %>% 
  filter(distsf < 1000) %>% 
  filter(car_driver > foot)
l_short = l_sf %>% 
  filter(distsf < 1000)
plot(l_drive_short[6])
```


```{r, eval=FALSE}
devtools::install_github("osmdatar/osmdata")
```


## Advanced challenge: download transport data with osmdata 

- You'll need to read-up on the osmdata vignette

```{r}
library(osmdata)
q = opq(bbox = "Leeds") %>% 
  add_feature(key = "railway", value = "station")
stations = osmdata_sf(q = q)
plot(l_sf[6])
plot(stations$osm_points, add = T)
```
