---
title: "Objects, functions and concepts for efficient R programming"
author: "Robin Lovelace, University of Leeds, \n `r Sys.Date()`"
date: "Slides: [rpubs.com/RobinLovelace](http://rpubs.com/RobinLovelace)"
output:
  ioslides_presentation:
    widescreen: true
    transition: slower
    logo: ../figures/its-logo-square.png
---


## R as a giant calculator

```{r, eval=FALSE}
5 * 5
```

```{r, eval=FALSE}
1 + 4 * 5
```

```{r, eval=FALSE}
4 * 5 ^ 2
```

```{r, eval=FALSE}
sin(90)
```

```{r, eval=FALSE}
sin(0.5 * pi)
```

```{r, eval=FALSE, echo=FALSE}
sin(90 * pi / 180)
```

## Objects

```{r}
a <- 1
b <- 2
c <- "c"
x_thingy <- 4
```

```{r, eval=FALSE}
a + b
a * b
a + c
a / x_thingy
```

## Adding and removing objects

```{r}
ls()
x <- x_thingy
rm(x_thingy)
x
ls()
```


## Harmonograph example

```{r, echo=FALSE, warning=FALSE}
t <- 1:1000 / 10
x <- sin(t) * exp(t * - 0.01)
y <- sin(pi / 2 + t / 3) * exp(t * -0.01)
plot(x, y, "lines")
```

## Practical 1: Getting used to RStudio and R

> - Open RStudio and have a look around
> - Create a new project
> - Create a new R Script: pass code to the console with `Ctl-Enter`
> - Use R as a calculator: what is:

$$
\pi * 9.15^2
$$

> - Explore each of the 'panes'
> - Find and write down some useful shortcuts (`Alt-Shift-K` on Windows/Linux)

# Basic R functions and behaviour

## Functions and objects

**In R:**

- Everything that exists is an object
- Everything that happens is a function

```{r}
# Assignment of x
x <- 5
x
```

```{r, eval=FALSE}
# A trick to print x
(x <- 5)
```

```{r, eval=FALSE}
(y <- x)
```

## Functions

```{r}
sin(x)
exp(x)
factorial(x)
sinx <- sin(x)
```

## R is vector based

```{r}
x <- c(1, 2, 5)
x
x^2
x + 2
```

```{r, eval=FALSE}
x + rev(x)
```

## The classic programming way: verbose

```{r}
x <- c(1, 2, 5)
for(i in x){
  print(i^2)
}
```

Creating a new vector based on `x`

```{r}
for(i in 1:length(x)){
  if(i == 1) x2 <- x[i]^2
  else x2 <- c(x2, x[i]^2)
}
x2
```

## Data types

R has a hierarchy of data classes, tending to the lowest:

- Binary
- Integer (numeric)
- Double (numeric)
- Character

## Examples of data types

```{r}
a <- TRUE
b <- 1:5
c <- pi
d <- "Hello Leeds"
```

```{r, eval=FALSE}
class(a)
class(b)
class(c)
class(d)
```

## Data type switching

```{r}
ab <- c(a, b)
ab
class(ab)
```


## Test on data types

```{r}
class(c(a, b))
class(c(a, c))
class(c(b, d))
```

## Sequences

```{r}
x <- 1:5
y <- 2:6
plot(x, y)
```

## Sequences with seq

```{r}
x <- seq(1,2, by = 0.2)
length(x)
x <- seq(1, 2, length.out = 5)
length(x)
```

## Practical: 45 minutes

Working through Chapter 20 of *R for Data Science*: http://r4ds.had.co.nz/vectors.html

- Work through 20.3

- Challenge: create an atomic vector of 3 characters and an `NA`

- Challenge: create an atomic vector of 5 numeric vectors

```{r, echo=FALSE}
x = 1:5
# class(x)
x = c(1:4, 4.9999999)
# class(x)
```

- Challenge: create a list containing logical, character and numeric vectors
    - What happens to the class if you *unlist* this list?

```{r, echo=FALSE, eval=FALSE}
l = list(TRUE, "yes", 5.5)
# unlist(l)
```

- Complete all the code examples in 20.4
    - Challenge: what are the outcomes of the following commands?:
    
```{r, eval=FALSE}
x = 1:10
x[c(1, NA)]
x[2 * 3]
x[-2 * 3]
```

## Data loading practical

- Run the code in `download-data.R`
- Check that the files have downloaded and unzipped OK
    - If so: you've used R as a file manager!
- Work through pages 4, 5 and 6 in the Creating-maps-in-R tutorial



