# 2 day course: R for Spatial Transport Data

Assumed prior knowledge:

- Some knowledge of tidyverse is highly recommended, e.g. already know or have read 

<!-- **It's in the Analysis directorate** -->


## Computing Requirements

- DfT will provide desktops with recent versions of R and RStudio installed, including:
  - RStudio 3.5.3 recommended (although any R 3.5 release should work): https://cran.r-project.org/
  - Will require up-to-date versions of **tidyverse** and **sf** packages, at a minimum
  - Other packages we'll use:
    - stats19
    - stplanr
    - tmap
    - devtools
    - The GitHub package spDataLarge, which can be installed as follows:
    
```{r, eval=FALSE}
devtools::install_github("Nowosad/spDataLarge")
```

- Own laptops allowed, provided necessary data is installed

- Data: all data will be either provided by packages or downloaded on the day (assuming good internet)
  - A test to check if data downloads work is accessing stats19 data (check this works):

```{r, eval=FALSE}
crashes = stats19::get_stats19(year = 2017, type = "accidents", ask = FALSE)
```

- A script to test set-up will be provided to test it all works

## Venue and course times

Venue: near DfT HQ at Horseferry road, TBC

Course times each day:

- 09:30 - 10:00 (set-up)

- 10:00 - 11:00

- break

- 11:30 - 12:30

- 12:30 - 13:30: lunch

- 13:30 - 15:00

- break

- 15:00 - 16:00

- 16:00 - 16:30 (Q&A)

Day 1:

Introduction to R/RStudio

- How to use R/RStudio effectively 
- Using packages: example with sf/ggplot2 (tbc)
- Spatial data analysis
- Visualising spatial datasets

Day 2:

- stats19 data analysis - with spatial/temporal analysis 
- OD data with stplanr
- Data and methods for assessing cycling potential
- Local route network analysis 
  - Download data from open roads: put on test scripts

